---
title: "LU Decomposition"
author: "Rishi Dey Chowdhury"
date: "7/30/2021"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set()
library(knitr)
```

Here we are going to implement an algorithm to find out the
$LU$-Decomposition of a Matrix Efficiently. We will then use it to solve
System of Equations. \#\#\# OBJECTIVE Implement the efficient version of
Crout's decomposition. Solve a system $AX=b$ by forward and backward
substitution.

### LU DECOMPOSITION and SOLVING SYSTEMS

Whenever we have a Matrix whose $LU$-Decomposition in possible, where
$L$ is lower triangular Matrix and $U$ is upper Triangular Matrix with
$1$ across it's diagonal. We can write $A$ as a factorization $A=LU$ and
we can solve the Matrix System $$Ax = b$$ $$LUx = b$$ Now, we can solve
2 Systems Quickly by Forward and Backward Substitution to get the
Solution that is $$Ly = b\ (Forward\ Substitution)$$
$$Ux = y\ (Backward\ Substitution)$$ Thus, we get the solution $x$ to
the original system.

### CODE & RESULTS

-   The following functions computes the $L$ and $U$ for a given matrix
    for which $LU$-Decomposition is possible. We at no step explicitly
    allocate separate space to store $L$ and $U$ as we are aware that
    the diagonal Matrix of $U$ are bound to be $1$. So, we simple use
    the space of the original Matrix to store $L$ and $U$ in it.

```{r}
# Computes the L matrix 
L <- function(A, j) {
  n <- nrow(A)
  tot <- rep(0, n-j+1)
  if (j > 1){
    for (k in 1:(j-1))
      tot <- tot + A[j:n, k] * A[k, j]
  }
  A[(j:n), j] <- A[(j:n), j] - tot
  return(A)
}

# Computes the U matrix
U <- function(A, i) {
  n <- nrow(A)
  tot <- rep(0, n-i)
  if (i > 1){
    for (k in 1:(i-1))
      tot <- tot + A[k, (i+1):n] * A[i, k]
  }
  if(A[i, i] == 0)
    stop("Singular Matrix Detected!")
  A[i, (i+1):n] <- (A[i, (i+1):n] - tot) / A[i, i]
  return(A)
}
```

-   We now use the above functions to compute the $LU$-Decomposition.

```{r}
# Computes the LU Decomposition
Compute.LU <- function(A) {
  if(nrow(A) != ncol(A))
    stop("Matrix is not square!")
  
  n <- nrow(A)
  
  for(i in 1:n) {
    A <- L(A, i)
    if(i < n)
      A <- U(A, i)
  }
  
  L <- matrix(rep(0, n^2), n, n)
  U <- diag(n)
  
  for(j in 1:n)
    L[j:n,j] <- A[j:n, j]
  
  for(i in 1:(n-1))
    U[i, (i+1):n] <- A[i, (i+1):n]
  
  return(list("A" = A, "L"=L, "U"=U))
}
```

Let's Try it out on some Matrix.

```{r}
A <- matrix(c(1,2,3,5,6,7,9,11,12), 3, 3)
kable(A)
```

The $LU$-Decomposition of the above Matrix.

```{r}
decomp <- Compute.LU(A)
kable(decomp$A, caption = "Packed Matrix")
kable(decomp$L, caption = "L")
kable(decomp$U, caption = "U")
```

The Specification of $L$ an $U$ being lower and upper triangular matrix
with $U$ having diagonal elements as $1$ is met. Let's test if $A = LU$.

```{r}
kable(decomp$L %*% decomp$U, caption = "LU")
```

Indeed we get back the original matrix. - Now, we are going to use the
above $LU$-Decomposition to obtain the solution for the System $Ax=b$.
The Following code implements this Forward Substitution and Backward
Substitution as discussed earlier.

```{r}
# Solve Linear System
Solve.LU <- function(A, b){
  LU <- Compute.LU(A)
  n <- nrow(A)
  
  # Forward Substitution
  y <- rep(0, n)
  if (LU$A[1,1] == 0) 
    stop("No Solution!")
  y[1] <- b[1] / LU$A[1,1] 
  for (i in 2:n) {
    if (LU$A[i,i] == 0)
      stop("No Solution!")
    y[i] <- (b[i] - sum(LU$A[i, 1:(i-1)]*y[1:(i-1)])) / LU$A[i,i] 
  }
  
  # Backward Substitution
  x <- rep(0, n)
  x[n] <- y[n]
  for (i in (n-1):1) {
    x[i] <- y[i] - sum(LU$A[i, (i+1):n]*x[(i+1):n])
  }
  
  return(list("LU"=LU, "x"=x))
}
```

Let's Test it on the Following System.

```{r}
A <- matrix(c(2,4,-6,4,3,7,-10,6,0,2,0,4,0,0,1,5), 4, 4)
kable(A, caption = "A")
kable(c(1,2,1,0), col.names = "b")
```

The solution is

```{r}
sol <- Solve.LU(A, c(1,2,1,0))
kable(sol$x, caption = "Soln.")
```

Let's check if $Ax=b$.

```{r}
kable(A %*% sol$x, col.names = "Ax")
```

Indeed we get $b$ and hence it is a solution to the System.
